Searching for standards of fairness in the transportation justice literature: A scoping review

### 2023-11-08 Manuscript/Manuscript.qmd
Incorporation of Ignacio comments into manuscript.

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all 
           #width = rep(10, each=50)
           ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.2, limits = c(0,.4), na.value = "#FCB5AC") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][1:5]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all %>% percent(accuracy = 0.1)), color = "white", size = 2.5) +
  geom_vline(xintercept=4.5, linewidth=1.5, color = "#191970") +
  geom_hline(yintercept = 43.5, linewidth=1.5, color="#191970")+
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank())+
  ggtext::geom_richtext(aes(x=5.385, y=23.5, label="Conceptualizations"), fill="grey", size=2) +
  ggtext::geom_richtext(aes(x=3, y=49, label="Standards"), angle = 270, fill="grey", size=2) 
# standards_conc_percentages_plot
# ggsave("standards_conc_percentages_plot.png")
```

```{r, warning=FALSE}
#the heatmap on top and table of stand/concept definitions on the bottom.
standards_conc_percentages_plot_and_table <- cowplot::plot_grid(
  gen_grob(totals_standards_table_flex, scaling="fixed", fit="auto"),
  standards_conc_percentages_plot,
  NULL,
  gen_grob(totals_conc_table_flex, scaling = "fixed", fit="auto"),
  nrow = 2, ncol=2, rel_heights = c(1.5,1), rel_widths = c(1,4) )
#standards_conc_percentages_plot_and_table
# ggsave("standards_conc_percentages_plot_and_table.png")
```

@@

- Accessibility:  
- Capabilities approach:  
- Equality:  
- Equity:  
- Fairness:  
- Justice:  
- Mobility:  
- Object of justice:  
- Policy:  
- Social exclusion:  
- Space-time convergence:  
- Subject of justice:   
- Sufficientarianism:  
- Theories of justice:  
- Utilitarianism:

Etc.,

@@

Coverage of the July 20 launch by the mainstream media was in many cases uncritical. "We're going to build a road to space so that our kids and their kids can build a future" Bezos declared, before adding "...we need to do that to solve the problems here on Earth" [@johnsonRoadToSpace2021]. Few reporters saw it fit to ask what problems the billionaire planned to solve on Earth, or what kind of future Bezos was trying to build, and for whose children. In other words, the billionaire was not confronted with questions about what did his trip do _for whom_ and _to whom_. A less dispassionate observer might have been excused for wondering (possibly aloud) about the basic _fairness_ of a man amassing a nigh unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his logistics empire.

@@

The rocket that took Bezos to the edge of space is a somewhat rare example of a transportation technology, a tool of space-time convergence. By enabling movement at very high speeds, rockets might—one day, at some indeterminate point in the future—prove essential to the expansion of the human species beyond our home planet. But in the present moment, the benefits of a private suborbital flight (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space) are for a few, whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that. The benefits of public transportation, a much more common transportation technology, are for most, but in many cases we have penny-pinched these systems while concentrating the costs on those who have less [@allenSuburbanization2020; @kaeorueanAnalysis2020]. In a plutocratic society, those with most can (and often do argue^[See the so-called [Techno-optimist Manifesto](https://a16z.com/the-techno-optimist-manifesto/)]) that this distribution of burdens and benefits is fair since both benefits and burdens are earned. The fact that the likes of Bezos do travel to space is proof that the likes of Bezos are due those trips. In a democratic society, the members of the collective might actually agree that large rewards (e.g., space flight) must be offered to highly qualified individuals (e.g., Bezos) to entice them to take important responsibilities (e.g., founding and leading Amazon). The fact that some must give up jobs because it takes them too long to reach them by public transportation would constitute proof that those people should have studied more, been earlier risers, worked harder [@SpurrWaiting2015; @greismanLittle2017]. Again, the members of the collective might think this a fine state of affairs, having come to this conclusion of their own accord or after being persuaded by billionaires. Or, contrariwise, the members of the collective might decide that this state of affairs is _unjust_: the _values_ of those in charge of defining what is "fair" matter.

@@

The objective of this report is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain. Much research has been devoted to the issues of _measuring _ equity in transportation, including (among many others) @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022. Further, there are multiple sources that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019]. Finally, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling; @doranPursuitCyclingEquity2021]. These inquiries are valuable to scholars, planning agencies, the public, and decision-makers alike, and the present review will tread similar, but not completely overlapping ground. In our estimation, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented in the transportation domain. In contrast, we do know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].

@@

Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution. Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy. In this, we aim to provide relief even, or particularly, to planners in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].

@@

**Retributive justice**. It is concerned with the retribution due to people who do wrongs.

**Reparative (or restorative) justice**. This form of justice relates to the proper way to correct or rectify past wrongs even if the wrongdoers can no longer be meted retribution.

**Procedural justice**. The main concern of this form of justice is to ensure that the views, opinions, and preferences of all stakeholders are properly accounted for when decisions that affect their lives are made.

**Distributive justice**. This is perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p. 2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.

It might be argued that all of the above are particular forms of distributive justice. Retributive justice, for example, is often concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or claiming tangibles like resources as burdens (e.g., fines). Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs. Procedural justice is the distribution of the benefits (e.g., the right to voice an opinion) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.

@@

The term "equity" tends to encompass various tools to understand the distribution of benefits and burdens of things among a population, with a particular emphasis on those with the least advantage and equality in their outcomes. In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020]. Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities. In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.

Fairness, in contrast to equity, is somewhat more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination, i.e., reparative justice). What kind of justice does fairness serve in each case? It depends on what was the rationale for seeking justice in the first place. Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.

@@

To recap the discussion of definitions so far, justice is an end. But to understand what that end is, we must clearly define standards of fairness. Equity analysis is tool to measure where the actual or perceived distribution of the burdens and benefits of the products of transportation systems stand with respect to the standard, in other words, instruments to see how close or far we are from a just situation.

@@

```{r plot-standards-concs-perc, warning=FALSE}
#| label: fig-fig6
#| fig-cap: "The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category."
standards_conc_percentages_plot_and_table
ggsave("standards_conc_percentages_plot_and_table.png")
```
<!--`r word_comment("Answering: what conceptualizations occur in papers with opportunity standards? what types of methods?", ".")`-->
We begin this section with our reflections about the way equity standards connect with conceptual and theoretical frameworks of justice. Broadly speaking, some trends emerge in terms of the methods used. **Opportunity** and **population** standards appear in the literature in similar proportions ( `r (stand_count$count[3] /sum(stand_count$count)) |> percent(accuracy=0.1)` vs. `r (stand_count$count[4] /sum(stand_count$count)) |> percent(accuracy=0.1)`), but frequently correspond to different conceptualizations of justice. Referring to @fig-fig6, much attention has been given to questions of **horizontal equity** and **spatial equity**, and over 40% of these papers propose standards that relate to **opportunities**. Within these papers, a frequent suggestion is to use standards that relate to travel impedance, which describes how fast, far, or costly it is to travel from one place to another. We think of this as a type of opportunity standard because travel impedance is essential to the measurement of accessibility (i.e., the ease of reaching opportunities). Some examples include the work of @chenImpactHighspeedRail2017, who use a standard of 4 hours or less by high-speed rail for municipalities to be considered "comfortably connected". In a similar manner, @yenisettyMeasuringAccessibilityVarious2020 assumes that a travel distance of less than 1,200 m is sufficient for a resident to interact with its transit system locally. And @shenMeasuringSpatialAccessibility2020 identify regions where populations cannot access hospitals within 1 hour by car - in other words, more than sixty minutes travel time is already seen as an unfair burden. Papers that suggest opportunity standards often employ disparity analysis through a variety of quantitative approaches. These include inequality measures such as the Gini coefficient and Lorenz curve, or poverty measures [@vanderveenOperationalizingIndicatorSufficient2020; @tiznadoaitkenRoleAccessibilityPublic2018]). There are also examples of spatial descriptive analysis as well as comparisons to benchmarks [e.g., equal supply to demand of public transit in a spatial unit as in @peungnumsaiGridBasedSpatialAnalysis2020] to determine which locations are spatially and horizontally (in)equitable. Another branch of quantitative research conceive the externalities of transportation system as trade-offs, and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies e.g., [@fengMulticriteriaEvaluationAccessibilitybased2014; @fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014]. Papers that focus primarily on **horizontal equity** and/or **spatial equity** seldom use exclusively qualitative methods.

A different way to approach standards is to see them from the lens of **population**, and when this is done, other methods are commonly used. Over 40% of papers that suggest **population**-based standards are founded on concepts of **well-being**. These papers typically ask *what is enough to lead a satisfactory life (as related to transportation)*, and the standards suggested include population benchmarks for comparison such as: questionnaires and relative comparisons to recommended physical activity per week [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @towneUsingWalkScore2016]. Summative per capita benchmarks include energy consumption for a 'decent living' [as suggested in @raoDecentLivingEmissions2012], and region-relative comparisons in health-related outcomes, such as premature mortality rates [@awuorInfluenceEnvironmentalHealth2019], spatial access to hospitals [@pereiraGeographicAccessCOVID192021], as well as spatial access to supermarkets, active-mode-usage, and Body Mass Index (BMI) [@murphySupermarketAccessTransport2017].
The majority of these papers use quantitative/mixed-methods to identify inequities in **wellbeing**, however, a minority do use exclusively qualitative methods to distill themes, as for example in the exploration of *perceived* quality of life by @berheAdaptationDissonanceQuality2014.

Papers that combine **population**- and **opportunity**-based standards often rely on **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6). Perhaps not surprisingly, these papers often include a combination of methods: questionnaires and other qualitative methods related to population standards and quantitative methods such as accessibility indices for opportunity standards are usually deployed. For instance, census data and the estimated proportion of households within some travel distance/time/availability to/of key destinations is used to identify a variety of social exclusionary situations [e.g., @mackettIncreasingAccessibilityCostEffectively2010; @chenExploringTravelCharacteristics2010; @danielsProposalAccessibilityPlanning2011; @sunPublicTransportAvailability2021; @sharmaPublicTransitAccessibility2021], the linkages between well-being and transport-related social exclusion [e.g., @delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].

Similar to papers interested in **wellbeing**, the majority of papers concerned with **social-exclusion** use quantitative/mixed-methods to identify areas, households, and/or populations at risk. They use a variety of methods to identify _where_ populations at risk may be located, such as clustering methods [@mohriClusteringMethodMeasuring2021]. Few of the papers that employ exclusively qualitative methods use survey data to study the willingness to travel, or the barriers to travel [e.g., @chenExploringTravelCharacteristics2010; @mehdizadehWalkingTimeSchool2017] or interviews/focus groups related to topics of unmet activity needs [@nordbakkeTransportUnmetActivity2015]. Travel that _did not_ take place is notoriously slippery to capture in standard surveys, and understanding whether this is a preference or the consequence of constraints is not straightforward either [e.g., @paezParticipation2012].

>>>>

```{r}
standards_conc_percentages_plot <- ggplot(totals_standards_conc, 
       aes(x=std_mcat, y=conc, fill= total_perc_in_all 
           #width = rep(10, each=50)
           ))+ 
  geom_tile() + 
  scale_fill_gradient2(low = "#AFAFED", mid= "#3F3FD4", high = "#191970", midpoint = 0.2, limits = c(0,.4), na.value = "#FCB5AC") +
  scale_x_discrete(limits=(totals_standards_conc$std_mcat)[order(totals_standards_conc$group)][1:5]) + 
  scale_y_discrete(limits=(totals_standards_conc$conc)[order(totals_standards_conc$group)] ) +
  geom_text(aes(label = total_perc_in_all %>% percent(accuracy = 0.1)), color = "white", size = 2.5) +
  geom_vline(xintercept=4.5, linewidth=1.5, color = "#191970") +
  geom_hline(yintercept = 43.5, linewidth=1.5, color="#191970")+
  theme_classic() +
  coord_flip()+
  theme(legend.position = "none",
        axis.ticks = element_blank(),
        axis.text.x=element_text(angle=20,hjust=1, colour = "black"),
        axis.text = element_text(size=7, colour = "black"),
        axis.title = element_blank())
#+ ggtext::geom_richtext(aes(x=5.385, y=23.5, label="Conceptualizations"), fill="grey", size=2) +
#  ggtext::geom_richtext(aes(x=3, y=49, label="Standards"), angle = 270, fill="grey", size=2) 
# standards_conc_percentages_plot
# ggsave("standards_conc_percentages_plot.png")
```

```{r, warning=FALSE}
#the heatmap on top and table of stand/concept definitions on the bottom.
standards_conc_percentages_plot_and_table <- cowplot::plot_grid(
  gen_grob(totals_standards_table_flex, scaling="fixed", fit="auto"),
  standards_conc_percentages_plot,
  NULL,
  gen_grob(totals_conc_table_flex, scaling = "fixed", fit="auto"),
  nrow = 2, ncol=2, rel_heights = c(0.8,0.7), rel_widths = c(0.8,4) )
#standards_conc_percentages_plot_and_table
# ggsave("standards_conc_percentages_plot_and_table.png")
```

@@

- **Fairness**: is achieved when a collective and individuals are treated equally and without bias as defined by their morals and values. (*maybe add this into the intro??*)
- **Equity**: the distribution of benefits and burdens of things among the population, with a particular emphasis on those with the least advantage and quality in their outcomes.
- **Justice**: mutually reinforced by the tools of fairness and equity, justice is a multi-dimensional political ideal attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2]. It ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440]. The process towards justice is fluid and can be identified through questions that answer what is the context, subject and object of justice, and what is the rationale for justice.
- **The temporal-spatial context of justice**: the answers to "When?" and "Where?"
- **The subjects of justice**: the answer to "Who?" is the focus of justice.
- **The objects of justice**: the answer to "What?" is the focus of justice; what are the things that characterise (in)justice. 
- **The methods for justice determination**: the answer to "How?" is justice defined and markers towards it tracked.
- **The rationale to justice**: the "Why?". Connects with the context, subject, object and methods used for justice determination.
- **Equity standards**: are a tool for answering this distributive question: how do we allocate burdens and benefits and to whom? Standards are s that when operationalized effectively define what is fair.
- **Accessibility**:  
- **Capabilities approach**:  
- **Equality**:  
- **Mobility**:  
- **Policy**:  
- **Social exclusion**:  
- **Space-time convergence**:  
- **Sufficientarianism**:  
- **Theories of justice**:  
- **Utilitarianism**:  
Etc.,

@@

Coverage of the July 20 launch by the mainstream media was in many cases uncritical. "We're going to build a road to space so that our kids and their kids can build a future" Bezos declared, before adding "...we need to do that to solve the problems here on Earth" [@johnsonRoadToSpace2021]. Few reporters saw it fit to ask what problems the billionaire planned to solve on Earth, or what kind of future Bezos was trying to build, and for whose children. In other words, the billionaire was not confronted with questions about what his trip did _for whom_ and _to whom_. A less dispassionate observer might have been excused for wondering (possibly aloud) about the basic _fairness_ of a man amassing a nigh unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his logistics empire.

@@

The rocket that took Bezos to the edge of space is a somewhat rare example of a transportation technology, a tool of space-time convergence. By enabling movement at very high speeds, rockets might—one day, at some indeterminate point in the future—prove essential to the expansion of the human species beyond our home planet. But in the present moment, the benefits of a private suborbital flight (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space) are for a few, whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that. The benefits of public transportation, a much more common transportation technology, are for most, but in many cases we have penny-pinched these systems while concentrating the costs on those who have less [@allenSuburbanization2020; @kaeorueanAnalysis2020]. In a plutocratic society, those with most can (and often do argue^[See the so-called [Techno-optimist Manifesto](https://a16z.com/the-techno-optimist-manifesto/)]) that this distribution of burdens and benefits is fair since both benefits and burdens are earned. The fact that the likes of Bezos do travel to space is proof that the likes of Bezos are due those trips. In a democratic society, the members of the collective might actually agree that large rewards (e.g., space flight) must be offered to highly qualified individuals (e.g., Bezos) to entice them to take important responsibilities (e.g., founding and leading Amazon). The fact that some must give up jobs because it takes them too long to reach them by public transportation would constitute proof that those people should have studied more, been earlier risers, worked harder [@SpurrWaiting2015; @greismanLittle2017]. Again, the members of the collective might this is a fine state of affairs, having come to this conclusion of their own accord or after being persuaded by billionaires. Or, contrariwise, the members of the collective might decide that this state of affairs is _unjust_: the _values_ of those in charge of defining what is "fair" matter.

@@

The objective of this report is to scan the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain. Much research has been devoted to the issues of _measuring_ equity in transportation, including (among many others) @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022. Further, there are multiple sources that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019]. Finally, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling; @doranPursuitCyclingEquity2021]. These inquiries are valuable to scholars, planning agencies, the public, and decision-makers alike, and the present review will tread similar, but not completely overlapping ground. In our estimation, there remains a gap in the literature in terms of understanding how standards for equity are developed and implemented in the transportation domain. In contrast, we do know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].

@@

Supporting the creation of (more) just transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; and 3) what values are embodied in normative statements about said distribution. Ultimately, this review aims to collate the existing academic knowledge on the matter, and present it in a manner useful to support the development and implementation of standards for equity in transportation planning and policy. In this, we aim to provide relief even, or particularly, to planners in those places where calls for justice are explicitly made through legislation^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].

@@

**Retributive justice** is concerned with the proportional retribution of wrongdoers with relation to legitimate punishers and the innocent [@sep-justice-retributive].

**Reparative (or restorative) justice** focuses on the reparation of caused harm; it centers 'reintegrative shaming' to restore victims, wrongdoers and the community according to moral values [@tylerRestorativeJusticeProcedural2006].

**Procedural justice** strives to ensure that the views and preferences of all stakeholders are fairly accounted for in the decision-making and inter-personal procedures affecting their lives and communities [@tylerRestorativeJusticeProcedural2006].

**Distributive justice** is perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p. 2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society are collected by different segments of a population.

It might be argued that all of the above touch on forms of distributive justice. Retributive justice, for example, could be framed as being concerned with the distribution of the benefits and burdens of being a member of society; the way it is usually achieved is by distributing intangibles of a society's moral values like "freedom" (e.g., of movement, of association) as benefits, and the distribution of tangible resources as burdens (e.g., fines as a punishment). Reparative justice could look like the distribution of benefits and burdens to redress past wrongs, for example by asking those who have benefited from said wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are still harmed by past wrongs. Procedural justice could be the distribution of the benefits (e.g., the right to voice an opinion as a recognized stakeholder in the process) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.

@@

The term "equity", as conceptualized alongside distributive justice, tends to encompass various tools to understand the distribution of benefits and burdens of things among a population; there is a particular emphasis on those with the least advantage and equality in their outcomes. In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020]. Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequities. In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, or else they may be more vulnerable to misuse and even abuse as tools of subjugation.

Fairness, in contrast to equity, is somewhat more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes, ranging from the banal (do not _knowingly_ discriminate), to the substantive (compensation for past discrimination within a societal context that recognizes harm was done, i.e., reparative justice). What kind of justice does fairness serve in each case? It depends on what was the rationale for seeking justice in the first place. Our reading of @martensFair2021 is that fairness is a yardstick that is best deployed _a priori_ than _a posteriori_, for doing the latter risks rationalizing the outcomes rather than driving them.

@@

To recap the discussion of definitions so far, justice is a goal of social progress. But to understand what that goal is, we must clearly define standards of fairness. Equity analysis is tool to measure where the actual or perceived distribution of the burdens and benefits of the products of transportation systems stand with respect to the standard, in other words, instruments to see how close or far we are from a just situation.

@@

```{r plot-standards-concs-perc, warning=FALSE}
#| label: fig-fig6
#| fig-cap: "The proportion of equity standards (vertical axis) within each type of equity conceptualization (horizontal axis) category."
standards_conc_percentages_plot_and_table
ggsave("standards_conc_percentages_plot_and_table.png")
```
<!--`r word_comment("Answering: what conceptualizations occur in papers with opportunity standards? what types of methods?", ".")`-->
We begin this section with our reflections about the way equity standards connect with conceptual and theoretical frameworks of justice. Broadly speaking, some trends emerge in terms of the methods used depending on the way equity is conceptualized and the type of standard suggested; See @fig-fig6 for definitions of of conceptualizations along the horizontal axis and types of standards along the vertical axis. These categories emerged as we reviewed the literature and we use them to discuss trends; they are by no means an exhaustive list of equity conceptualizations and types of standards.

Referring to @fig-fig6, **Opportunity** and **population** are the most commonly suggested standards in the literature reviewed; they appear in the literature in similar proportions ( `r (stand_count$count[3] /sum(stand_count$count)) |> percent(accuracy=0.1)` vs. `r (stand_count$count[4] /sum(stand_count$count)) |> percent(accuracy=0.1)`), but frequently correspond to different conceptualizations of equity. Papers suggesting **opportunity standards** often address questions of **horizontal equity** and **spatial equity** and suggest standards that relate to travel impedance; these standards describes how fast, far, or costly it is to travel from one place to another. We think of this as a type of opportunity standard because travel impedance is essential to the measurement of accessibility (i.e., the ease of reaching opportunities). Some examples include the work of @chenImpactHighspeedRail2017, who uses a standard of 4 hours or less by high-speed rail for municipalities to be considered "comfortably connected". In a similar manner, @yenisettyMeasuringAccessibilityVarious2020 assumes that a travel distance of less than 1,200 m is sufficient for a resident to interact with its transit system locally. And @shenMeasuringSpatialAccessibility2020 identify regions where populations cannot access hospitals within 1 hour by car - in other words, more than sixty minutes travel time is already seen as an unfair burden. These papers uses these standards as a benchmark to indicate what areas are better or worse-off in terms of travel impedance; we can interpret them to be a type of _difference_ threshold.

Papers that suggest these **opportunity** standards sometimes also employ disparity analysis through a variety of quantitative approaches to discuss the fairness in distribution. These include inequality measures such as the Gini coefficient and Lorenz curve, or poverty measures e.g., [@vanderveenOperationalizingIndicatorSufficient2020; @tiznadoaitkenRoleAccessibilityPublic2018]). They can also include spatial descriptive analysis as well as comparisons to service benchmarks [e.g., equal supply to demand of public transit in a spatial unit as in @peungnumsaiGridBasedSpatialAnalysis2020] to determine which locations are spatially and horizontally (in)equitable. Another branch of opportunity-standard-suggesting papers conceive the externalities of transportation system as trade-offs, and aim to maximize transport-related benefits (i.e., time savings, emissions reductions, congestion reductions, user fares) through optimization/location-allocation methodologies e.g., [@fengMulticriteriaEvaluationAccessibilitybased2014; @fakhrmoosaviIncorporatingTravelTime2021; @zhengAreabasedEquitablePricing2020; @wismadiSpatialPreferenceModelling2014]. In the reviewed papers that focus primarily on **horizontal equity** and/or **spatial equity** seldom use exclusively qualitative methods and operationalise predominately quantitative methods.

A different way to approach standards is to see them from the lens of **population**, and when this is done, other methods are commonly used, often depending on how equity is conceptualised. Papers that suggest **population**-based standards are often founded on concepts of **well-being**: they typically ask *what is enough to lead a satisfactory life (as related to transportation)?*. Standards suggested include: questionnaires and relative comparisons to recommended physical activity per week e.g., [@adlakhaMindGapGender2020; @auchinclossDesignBaselineDescription2020; @mccormackRelationshipClusteranalysisDerived2012; @kimNeighborhoodEnvironmentWalkability2016; @towneUsingWalkScore2016], summative per capita benchmarks (e.g., such as energy consumption for a 'decent living' as suggested in @raoDecentLivingEmissions2012), region-relative comparisons in health-related outcomes (e.g., premature mortality rates [@awuorInfluenceEnvironmentalHealth2019]), spatial access benchmarks to hospitals e.g., [@pereiraGeographicAccessCOVID192021], as well as spatial access to supermarkets, active-mode-usage, and Body Mass Index (BMI) e.g., [@murphySupermarketAccessTransport2017].
The majority of these papers use quantitative/mixed-methods to identify inequities in **wellbeing**. However, a minority do use exclusively qualitative methods to distill themes, as for example in the exploration of perceived quality of life by @berheAdaptationDissonanceQuality2014.

Papers that combine **population**- and **opportunity**-based standards often rely on **vertical equity** and **transport-related social exclusion** (note the similar proportions in these standards in @fig-fig6). Perhaps not surprisingly, these papers commonly use a combination of methods: questionnaires and other qualitative methods related to population standards and quantitative methods such as accessibility indices for opportunity standards are usually deployed. For instance, census data and the estimated proportion of households within some travel distance/time/availability to/of key destinations is used to identify a variety of social exclusionary situations e.g., [@mackettIncreasingAccessibilityCostEffectively2010; @chenExploringTravelCharacteristics2010; @danielsProposalAccessibilityPlanning2011; @sunPublicTransportAvailability2021; @sharmaPublicTransitAccessibility2021], the linkages between well-being and transport-related social exclusion e.g.,[@delboscTransportProblemsThat2011; @churchillTransportPovertySubjective2019], areas more likely to experience transport poverty [@allenSizingTransportPoverty2019a], food deserts [@mckeyCrowdsourcedMappingHealthy2020], or transport-related energy poverty e.g., [@robinsonDoubleEnergyVulnerability2020; @berryInvestigatingFuelPoverty2016; @berryDistributionalEffectsCarbon2019].

Similar to papers interested in **wellbeing**, the majority of papers concerned with **social-exclusion** use quantitative/mixed-methods to identify areas, households, and/or populations at risk. They use a variety of methods to identify _where_ populations at risk may be located, such as clustering methods [@mohriClusteringMethodMeasuring2021]. Some papers that exclusively use qualitative methods operationalise survey data to study the willingness/barriers to travel e.g., [@chenExploringTravelCharacteristics2010; @mehdizadehWalkingTimeSchool2017] or conducts interviews on topics of unmet activity needs [@nordbakkeTransportUnmetActivity2015]. Travel that _did not_ take place is notoriously slippery to capture in standard surveys, and understanding whether this is a preference or the consequence of constraints is not straightforward [e.g., @paezParticipation2012].
